NumPy---Array Programming in Python

### 2020-01-25 body.tex
Adopt review comments from Stefan in interop sectio

Users predominantly interact with NumPy arrays using {\em indexing} (to access
subarrays or individual elements), {\em operators} (e.g., $+$, $-$, $\times$,
and, for matrix multiplication, $@$), as well as {\em array-aware functions};
together, these provide an easily readable, expressive, high-level API for
array programming.

Various libraries build on and extend this functionality to further enrich the
ecosystem of available tools.  For example, SciPy adds various mathematical,
scientific, and engineering routines; Pandas provides indexed arrays, known as
dataframes; xarray implements arrays with labeled axes; and Matplotlib
generates publication-ready figures and visualizations.  Employing these
low-level array-aware functions are widely used technique-specific libraries,
including scikit-learn (machine learning), scikit-image (image processing),
statsmodels (statistics),  networkx (graphs and complex networks)
\cite{SciPyProceedings_11}, and Napari (volumetric image visualization).

Finally, domain-specific libraries and applications are built on this diverse
foundation of tooling; these include Astropy (astronomy) \cite{astropy:2013,
astropy:2018}, Biopython (computational biology and bioinformatics) \cite{cock2009biopython},
NIPY (neuroimaging) \cite{millman2007analysis}, and Pangeo (earth science) \cite{2018EGUGA..2012146H}.

Exposing array programming primitives, as well as the surrounding ecosystem of
tools, in Python---an interpreted language---creates an ideal environment for
interactive, exploratory data analysis where users may iteratively inspect,
manipulate, and visualize their data.

IPython \cite{perez2007ipython}

Jupyter \cite{Kluyver:2016aa}

\section*{The proliferation of arrays}

\fixme{Add references somewhere to Zarr, https://arrow.apache.org/,
https://parquet.apache.org/,
https://github.com/ruby-numo/numo-narray/wiki/Comparison-with-Numpy}

@@

This includes popular tensor computation libraries such as
CuPy\footnote{\url{https://cupy.chainer.org/}},
JAX\footnote{\url{https://jax.readthedocs.io/en/latest/jax.numpy.html}}, and
Apache MXNet\footnote{\url{https://numpy.mxnet.io/}}.
PyTorch\footnote{\url{https://pytorch.org/tutorials/beginner/blitz/tensor\_tutorial.html}}
and
Tensorflow\footnote{\url{https://www.tensorflow.org/tutorials/customization/basics}}
provide tensor APIs with NumPy-inspired semantics.

@@

To facilitate \emph{interoperability}, the NumPy team has designed
``protocols'' (or contracts of operation), that allow for these arrays to be
passed to NumPy functions (Fig.~\ref{fig:duck-arrays}).
NumPy, in turn, dispatches operations to the originating library, as required.
% FIXME: expand to 3 or more sentences discussing current efforts and
% issues with those, how did adoption go, what is being done next
%\url{https://numpy.org/neps/nep-0037-array-module.html}
Versions of these protocols have been successfully deployed.
% Ralf: Could you provide some proof of effectiveness, e.g., "This has made it
% possible to create distributed GPU arrays, enabling .... [find a reference to
% work by Peter Entschev]".

@@

NumPy was initially developed by students, faculty, and researchers to
provide a modern array programming library for Python.  These
user-developers frequently had to write code from scratch to solve
their or their colleagues' problems---often in low-level languages
that preceded Python, like Fortran \cite{dongarra2008netlib} and C.
To them, the advantages of an interactive, high-level array library
were evident. Their design of this new tool was informed by their
experiences with powerful interactive programming languages for
scientific computing such as APL \cite{iverson1962programming} and
Yorick \cite{munro1995using}, as well as commercial languages and
environments like IDL and Matlab.

NumPy has subsequently seen prolific adoption, with millions of users,
first from research and later also industry.
It spurred the development of the larger scientific Python ecosystem and
heralded the current era of wide-spread use of Python for scientific computing.
The reasons for its success are complex, but include the fact NumPy
(a) is built on top of Python, (b) has a strong culture of collaboration
and good software development practices, and (c) provides one of the
best array programming environments for scientific data analysis.

\emph{Python} is an open source, general-purpose, interpreted programming language
well-suited to standard programming tasks such as cleaning data,
interacting with web resources, and parsing text.
Adding fast array operations and linear algebra allows scientist to do all
their work within a single language---and one that has the advantage of
being famously easy to learn and teach.
In fact, Python's adoption as
a primary learning language in many universities (e.g., many lower division
UC Berkeley courses use Python including Foundations of Data Science,
The Beauty and Joy of Computing,
Introduction to Computational Thinking with Data,
and The Structure and Interpretation of Computer Programs) has
further popularized NumPy as a tool of choice for modern data science.

Even though NumPy is not part of Python's standard library,
we have a health relationship to the core development team.
\fixme{we've gotten a number of features: slicing syntax, implicit tuples,
complex numbers, buffer protocol, matrix multiplication operator}
\fixme{but we benefit from separate governance}

@@

The \emph{array programming} paradigm originated with APL in the 50s and 60s.
It is characterized by working interactively with array data using
vectorized operations.
Its draw for scientist is captured by the following description of
how APL users work \cite{kromberg2007apl}:
\begin{quote}
Users of APL tend to ``live in their data''. Interactive interpreters allow them
to inspect, almost feel their  way forward, discovering successful snippets
of code through experimentation and collecting them
into imperative or functional programs according to taste.
\end{quote}
Scientist exploring data at the IPython command prompt with the
whole scientific Python ecosystem available to them often work in
this way.  This type of interactive scientific exploration is
increasingly taking place in Jupyter notebooks.

NumPy combines the performance of C, the expressive power of array
programming, and the Python programming language in a mature
well-tested and well-documented community developed library.
It builds on decades of work refining the array programming
syntax from APL through Yorick and others.
It also made it easy for other libraries to develop fast and
memory-efficient compiled code, usually in C or Fortran, that could manipulate
these arrays and pass them back to Python.
Popular libraries increased the user community and that
in turn led to new libraries adding additional functionality
to the scientific Python ecosystem in a virtuous cycle.

>>>>

The brevity and expressive power of NumPy's array programming syntax combined with 
an advanced interactive environment like IPython \cite{perez2007ipython}
or Jupyter \cite{Kluyver:2016aa} makes interactive data exploration fluid and simple.
It is common for NumPy users to work interactively with their data,
discovering through experimentation bits of code that they stitch together into
imperative or functional programs.
The scientific Python ecosystem (Fig.~\ref{fig:ecosystem}) builds on top of
this solid foundation with core (e.g., SciPy and Matplotlib) libraries,
widely used technique-specific
libraries\cite{pedregosa2011scikit,vanderwalt2014scikit,SciPyProceedings_11},
and several domain-specific projects and libraries
\cite{astropy:2013,astropy:2018,cock2009biopython,millman2007analysis,2018EGUGA..2012146H}.
Exposing these simple array programming primitives, as well as the surrounding ecosystem of
tools, in Python---an interpreted, general purpose language---creates an ideal environment for
interactive, exploratory data analysis where users may iteratively inspect,
manipulate, and visualize their data.
This rich environment has made the scientific Python ecosystem popular for scientific
research.

\begin{figure}
  \centering
  \includegraphics[width=.48\textwidth]{static/ecosystem}
  \caption{\textbf{Scientific Python Ecosystem.}
   This figure highlights some of the essential tools and libraries in the vast
   scientific Python ecosystem.  Most of the ecosystem builds on NumPy to
   provide a rich and powerful collection of array-aware functionality.
  }
  \label{fig:ecosystem}
\end{figure}

For example, the \code{eht-imaging} library developed by the Event Horizon Telescope 
collaboration relies on many components of the scientific Python ecosystem.
NumPy arrays are used to store and manipulate numerical data at every step
in the processing chain: from raw data through calibration and image 
reconstruction.
Scipy is used to provide tools used in general image processing such as 
filtering and image alignment, while scikit-image provides higher-level
functionality such as edge filters and Hough transforms.
SciPy provides fundamental algorithms for scientific computing on arrays
including many mathematical, scientific, and engineering routines.
SciPy and NumPy are closely related and were developed together often
by the same people.
scikit-image is a popular technique-specific library,
which vastly extends the basic image processing tools provided by SciPy.
Other components of Scipy are used throughout the library, such as the
\code{scipy.optimize} package for handling general optimization tasks.
NetworkX \cite{SciPyProceedings_11} (the standard package for complex
network analysis) is used for analysis of consistency among image
comparisons.
Astropy \cite{astropy:2013, astropy:2018} (a community-developed astronomy
library) is used throughout the library,
primarily for I/O from standard astronomical file formats and time/coordinate
transformations.
Matplotlib (the core library for generating publication-ready figures
and visualizations) is used for visualizing data throughout the analysis pipeline,
including the generation of the final image of the black hole.
% N.B. - most of this comes from introspecting the Image class defined in
% ehtim/image.py and grepping the dependencies in requirements.txt

Tools for analyzing data from gravitational wave observatories such as LIGO
and Virgo are provided by the \code{pycbc} package, which was used in the first
detection of gravitational waves \cite{abbott2016observation} and in 
on-going analysis of data from LIGO and Virgo.
\code{pycbc} make extensive use of the scientific Python ecosystem.
The time-series data from the interferometers are stored in NumPy arrays,
while the \code{scipy.signal} is used to construct and apply filters to the
data.
Matplotlib is used to visualize data at all points along the analysis chain
including raw data from the instrumentation and the time-frequency 
visualization of the ``chirps'' that result from mergers of compact binaries
such as black holes\cite{abbott2016observation} and
neutron stars\cite{abbott2017multi}.

The rich ecosystem of tools building on NumPy and the larger community
of scientists and programmers using those tools mean that NumPy has a
responsibility to maintain considerable stability.
This need creates interesting challenges for NumPy development as the
computational environment scientists are analyzing their data on continues
to evolve and change.

\section*{Proliferation of arrays}

%\fixme{Add references somewhere to Zarr, https://arrow.apache.org/,
%https://parquet.apache.org/,
%https://github.com/ruby-numo/numo-narray/wiki/Comparison-with-Numpy}

% pandas, xarray, ...

@@

This includes popular tensor computation libraries such as
CuPy, JAX, and Apache MXNet.
PyTorch and Tensorflow provide tensor APIs with NumPy-inspired semantics.
% \footnote{\url{https://cupy.chainer.org/}}
% \footnote{\url{https://jax.readthedocs.io/en/latest/jax.numpy.html}}
% \footnote{\url{https://numpy.mxnet.io/}}
% \footnote{\url{https://pytorch.org/tutorials/beginner/blitz/tensor\_tutorial.html}}
% \footnote{\url{https://www.tensorflow.org/tutorials/customization/basics}}

@@

To facilitate \emph{interoperability}, the NumPy team has designed
``protocols'' (or contracts of operation), that allow for these arrays to be
passed to NumPy functions (Fig.~\ref{fig:duck-arrays}).
NumPy, in turn, dispatches operations to the originating library, as required.
% FIXME: expand to 3 or more sentences discussing current efforts and
% issues with those, how did adoption go, what is being done next
%\url{https://numpy.org/neps/nep-0037-array-module.html}
Versions of these protocols have been successfully deployed.
% Ralf: Could you provide some proof of effectiveness, e.g., "This has made it
% possible to create distributed GPU arrays, enabling .... [find a reference to
% work by Peter Entschev]".
% Below an attempt to start on this:
Thanks to these developments users of Dask and NumPy arrays are now capable
to scale their computation to the GPU using Dasks CuPy integration with
only minimal code changes.
However, general adoption has not yet been achieved mainly for compatibility
concerns.
Thus, we are actively addressing these concerns by revising and extending
these protocols with the aim of wide interoperability that spans the whole
scientific Python ecosystem up to and including domain libraries.

@@

NumPy was initially developed by students, faculty, and researchers to
provide an advanced, open source array programming library for Python,
which was free to use and unencumbered by license servers, dongles, and the like.
There was a sense of building something
impactful together, for the benefit of many others.  Participating in
such an endeavor, within a welcoming community of like-minded
indivuals, held a powerful attraction for many early contributors.

These user-developers frequently had to write code from scratch to solve
their or their colleagues' problems---often in low-level languages
that preceded Python, like Fortran \cite{dongarra2008netlib} and C.
To them, the advantages of an interactive, high-level array library
were evident. The design of this new tool was informed by their
experiences with powerful interactive programming languages for
scientific computing such as APL \cite{iverson1962programming} and
Yorick \cite{munro1995using}, as well as commercial languages and
environments like IDL and Matlab.

NumPy has subsequently seen prolific adoption, with millions of users,
first from research and later also industry.
It spurred the development of the larger scientific Python ecosystem and
heralded the current era of wide-spread use of Python for scientific computing.
The reasons for its success are varied and complex, but include the fact that
NumPy (a) is built on top of Python,
(b) has a strong culture of collaboration and good software development practices, and
(c) provides a cutting-edge array programming environment for scientific data analysis.

\emph{Python} is an open source, general-purpose, interpreted programming language
well-suited to standard programming tasks such as cleaning data,
interacting with web resources, and parsing text.
Adding fast array operations and linear algebra allows scientist to do all
their work within a single language---and one that has the advantage of
being famously easy to learn and teach.
In fact, Python's adoption as a primary learning language in universities
(e.g., many lower division UC Berkeley courses use Python including
Foundations of Data Science,
The Beauty and Joy of Computing,
Introduction to Computational Thinking with Data,
and The Structure and Interpretation of Computer Programs) has
further popularized NumPy as a tool of choice for modern data science.

Even though NumPy is not part of Python's standard library,
we benefit from a good relationship with the Python developers.
Over the years, the Python language has added new features and
special syntax so that NumPy would have more succinct and 
easy to read array notation.
Since we are not part of the standard library, we are able to
dictate our own release policies and development patterns.
%\fixme{we've gotten a number of features: slicing syntax, implicit tuples,
%complex numbers, buffer protocol, matrix multiplication operator}
%\fixme{but we benefit from separate governance}

@@

NumPy combines the expressive power of \emph{array programming}, 
the performance of C, and the Python programming language in a mature,
well-tested, well-documented, and community developed library.
It builds on decades of work refining the array programming
syntax from APL through Yorick and others.
It also made it easy for other libraries to develop fast and
memory-efficient compiled code, usually in C or Fortran, that could manipulate
these arrays and pass them back to Python.
Popular array-aware libraries increased the user community and that
in turn led to more libraries adding new functionality
to the scientific Python ecosystem.
